---
title: "Polygon preparation"
author: "Robert Schlegel"
date: "2019-05-13"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

```{r global_options, include = FALSE}
knitr::opts_chunk$set(fig.width = 8, fig.align = 'center',
                      echo = TRUE, warning = FALSE, message = FALSE, 
                      eval = TRUE, tidy = FALSE)
```

## Introduction

This markdown file contains all of the code used to prepare the polygons that are used to define the different regions in the Northwest Atlantic. These different regions will then have their SST pixels spatially averaged to create a single time series per region. This is done so that the MHW detection algorithm may then be run on these individual time series to represent a general representation of the SST in those regions, rather than running the algorithm on each pixel individually, which would introduce a host of problems.

```{r}
# Packages used in this vignette
library(tidyverse)
library(R.matlab)
library(marmap)
```

## Coastal region polygons

The first step in this analysis is to broadly define the coastal regions based on previous research into thermally relevant boundaries. We have chosen to use a paper by Richaud et al 2016 to do this (https://www.sciencedirect.com/science/article/pii/S0278434316303181#f0010). Being the kind-hearted man that he is, Benjamin forwarded us the polygons (Richaud et al. 2016; Figure 2) from this work as a MATLAB file so we must first open that up and convert it to an R format for further use.

```{r mat-R, eval=FALSE}
# Load the file
NWA_polygons <- R.matlab::readMat("data/boundaries.mat")

# Remove index list items and attributes
NWA_polygons[grepl("[.]",names(NWA_polygons))] <- NULL
# attributes(NWA_polygons) <- NULL

# Function for neatly converting list items into a dataframe
# vec <- NWA_polygons[1]
mat_col <- function(vec){
  df <- as.data.frame(vec)
  df$region <- substr(colnames(df)[1], 2, nchar(colnames(df)[1]))
  colnames(df)[1] <- strtrim(colnames(df)[1], 1)
  df <- df[c(2,1)]
  return(df)
}

# Create multiple smaller data.frames
coords_1 <- cbind(mat_col(NWA_polygons[1]), mat_col(NWA_polygons[2])[2])
coords_2 <- cbind(mat_col(NWA_polygons[3]), mat_col(NWA_polygons[4])[2])
coords_3 <- cbind(mat_col(NWA_polygons[5]), mat_col(NWA_polygons[6])[2])
coords_4 <- cbind(mat_col(NWA_polygons[7]), mat_col(NWA_polygons[8])[2])
coords_5 <- cbind(mat_col(NWA_polygons[9]), mat_col(NWA_polygons[10])[2])
coords_6 <- cbind(mat_col(NWA_polygons[11]), mat_col(NWA_polygons[12])[2])

# Combine them into one full dataframe and save
NWA_coords <- rbind(coords_1, coords_2, coords_3, coords_4, coords_5, coords_6)
saveRDS(NWA_coords, "data/NWA_coords.Rda")
```

With our polygons now switched over from MATLAB to R we now want to visualise them to ensure that everything has gone smoothly.

```{r poly-vis}
# Load polygon coordinates
NWA_coords <- readRDS("data/NWA_coords.Rda")

# The base map
map_base <- ggplot2::fortify(maps::map(fill = TRUE, col = "grey80", plot = FALSE)) %>%
  dplyr::rename(lon = long) %>%
  mutate(group = ifelse(lon > 180, group+9999, group),
         lon = ifelse(lon > 180, lon-360, lon)) %>% 
  select(-region, -subregion)

# Quick map
NWA_coords_plot <- ggplot(data = NWA_coords, aes(x = x, y = y)) +
  geom_polygon(data = map_base, aes(x = lon, y = lat, group = group), show.legend = F) +
  geom_polygon(aes(colour = region, fill = region), size = 1.5, alpha = 0.2) +
  coord_cartesian(xlim = c(min(NWA_coords$x)-1, max(NWA_coords$x)+1),
                  ylim = c(min(NWA_coords$y)-1, max(NWA_coords$y)+1)) +
  labs(x = NULL, y = NULL, colour = "Region", fill = "Region") +
  theme(legend.position = "bottom")
ggsave(NWA_coords_plot, filename = "output/NWA_coords_plot.pdf", height = 5, width = 6)

# Visualise
NWA_coords_plot
```

The region abbreviations are: "gm" for Gulf of Maine, "gls" for Gulf of St. Lawrence, "ls" for Labrador Shelf, "mab" for Mid-Atlantic Bight, "nfs" for Newfoundland Shelf and "ss" for Scotian Shelf.

Before we move on, we'll do a bit of house keeping to establish a consistent study area for this project based on our polygons. We'll simply extend the study area by 1 degree of longitude and latitude from the furthest edges of the polygons, as seen in the figure above.

```{r study-area-coords}
# Set the max/min lon/at values
lon_min <- round(min(NWA_coords$x)-1, 2)
lon_max <- round(max(NWA_coords$x)+1, 2)
lat_min <- round(min(NWA_coords$y)-1, 2)
lat_max <- round(max(NWA_coords$y)+1, 2)

# Combine and save
NWA_corners <- c(lon_min, lon_max, lat_min, lat_max)
saveRDS(NWA_corners, file = "data/NWA_corners.Rda")
```


### Cabot Strait

It was decided that because we are interested in the geogrophy of the regions, and not just their temperature regimes, the Cabot Strait needed to be defined apart from the Gulf of St. Lawrence region. To do this we will simply snip the "gsl" polygon into two pieces at its narrowest point.

```{r cabot-strait-1}
# Extract the gsl region only
gsl_sub <- NWA_coords[NWA_coords$region == "gsl",]

# Add a simple integer column for ease of plotting
gsl_sub$row_count <- 1:nrow(gsl_sub)

ggplot(data = gsl_sub, aes(x = x, y = y)) +
  geom_polygon(aes(fill = region)) +
  geom_label(aes(label = row_count))
```

It appears from the crude figure above that we should pinch the polygon off into two separate shapes at row 6 and 10.

```{r cabot-strait-2}
# Create smaller gsl polygon
gsl_new <- NWA_coords[NWA_coords$region == "gsl",] %>% 
  slice(-c(7:9))

# Create new cbs (Cabot Strait) polygon
cbs <- NWA_coords[NWA_coords$region == "gsl",] %>% 
  slice(6:10) %>% 
  mutate(region = "cbs")

# Attach the new polygons to the original polygons
NWA_coords_cabot <- NWA_coords %>% 
  filter(region != "gsl") %>% 
  rbind(., gsl_new, cbs)
saveRDS(NWA_coords_cabot, "data/NWA_coords_cabot.Rda")

# Plot the new areas to ensure everything worked
NWA_coords_cabot_plot <- ggplot(data = NWA_coords_cabot, aes(x = x, y = y)) +
  geom_polygon(data = map_base, aes(x = lon, y = lat, group = group), show.legend = F) +
  geom_polygon(aes(colour = region, fill = region), size = 1.5, alpha = 0.2) +
  coord_cartesian(xlim = c(min(NWA_coords$x)-1, max(NWA_coords$x)+1),
                  ylim = c(min(NWA_coords$y)-1, max(NWA_coords$y)+1)) +
  labs(x = NULL, y = NULL, colour = "Region", fill = "Region") +
  theme(legend.position = "bottom")
ggsave(NWA_coords_cabot_plot, filename = "output/NWA_coords_cabot_plot.pdf", height = 5, width = 6)

# Visualise
NWA_coords_cabot_plot
```

Everything is looking good, but we may want to divide the Gulf of Maine (gm) into two polygons as well. This would make the Bay of Fundy in it's own region. FOr now however we will move on to the next step, which is dividing the current polygons by bathymetry.


## Bathymetry polygons

These regions are an excellent start, but because the aim of this research project is to determine the primary drivers of MHWs in the Northwest Atlantic we want to subset the regions not only byu their thermal characteristics, but also by their geophysical characteristics, which in this instance is most immediately available as bathymetry. The rationale for this is that surface forcing as a driver of anaomolously warm seawater is likely to be more prevalent in shallow waters than in shelf waters or deeper. To this end we will divide the regions into two sub-regions each: 0 -- 50 m depth, 51 -- 200 m depth. Unless otherwise specified we will not be looking at waters further than the continental shelf. This is because the usefulness of this work will largely be constrained by it's applications to coastal ecosystems and fisheries. Therefore we are less interested in understanding the drivers of synoptic scale and greater open ocean events.

Before we can divide up our polygons by bathymetry, we must first download said bathymetry data. We will use here the NOAA data as it is (to my knowledge) the most conveient bathymetry data available directly through R.

```{r bathy-download, eval=FALSE}
# Donwload NOAA bathy data at highest available resolution
NWA_bathy <- as.xyz(getNOAA.bathy(lon1 = NWA_corners[1], lon2 = NWA_corners[2], 
                                  lat1 = NWA_corners[3], lat2 = NWA_corners[4], resolution = 1))
colnames(NWA_bathy) <- c("lon", "lat", "depth")
saveRDS(NWA_bathy, file = "data/NWA_bathy.Rda")
```

With our high res bathymetry data downloaded we will now go about finding the 50 and 200 metre isobaths. These lines will be used to define the edges of our sub-regions. The naming conventions we will use will be

```{r isobaths}
# Load bathymetry data
NWA_bathy <- readRDS("data/NWA_bathy.Rda")


```


```{r poly-sub-region}
# Create depth polygons inside of region polygons

# Save new sub-region polygons

```


With our new sub-regions created let's take a look at them.

```{r poly-sub-plot}

```

Now that we have our sub-regions defined we may go about creating SST time series for each. This work is continued in the vignette 'SST preparation'.

