---
title: "Self-organising map (SOM) analysis"
author: "Robert Schlegel"
date: "2019-05-13"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

```{r global_options, include = FALSE}
knitr::opts_chunk$set(fig.width = 8, fig.align = 'center',
                      echo = TRUE, warning = FALSE, message = FALSE, 
                      eval = TRUE, tidy = FALSE)
```


## Introduction

This markdown file will contain the code used to perform the self-organising map (SOM) analysis on the prepared variable data as seen in the [Variable preparation](https://robwschlegel.github.io/MHWNWA/var-prep.html) vignette.

```{r libraries}
# Packages used in this vignette
library(tidyverse) # Base suite of functions
library(ncdf4) # For opening and working with NetCDF files
library(lubridate) # For convenient date manipulation

# Set number of cores
doMC::registerDoMC(cores = 50)

# Disable scientific notation for numeric values
  # I just find it annoying
options(scipen = 999)
```

## More

It would be interesting to see if the SOM outputs differ in any meaningful wayss when only data from the first half of the study time period are used compared against the second half.
