---
title: "SST preparation"
author: "Robert Schlegel"
date: "2019-07-24"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
csl: FMars.csl
bibliography: MHWNWA.bib
---

```{r global_options, include = FALSE}
knitr::opts_chunk$set(fig.width = 8, fig.align = 'center',
                      echo = TRUE, warning = FALSE, message = FALSE, 
                      eval = TRUE, tidy = FALSE)
```

## Introduction

Building on the work performed in the [Polygon preparation](https://robwschlegel.github.io/MHWNWA/polygon-prep.html) vignette, we will now create grouped SST time series for the regions in our study area. We will do this by finding which NOAA OISST pixels fall within each of the region polygons. Also note that throughout this vignette (and this entire project) we will use the climatology period of 1993 -- 2018 as this is the shortest time limiting us by the data products used for the various abiotic variables needed for the SOM.

```{r libraries}
# Packages used in this vignette
library(jsonlite, lib.loc = "../R-packages/")
library(tidyverse) # Base suite of functions
library(heatwaveR, lib.loc = "../R-packages/") # For detecting MHWs
# cat(paste0("heatwaveR version = ", packageDescription("heatwaveR")$Version))
library(FNN) # For fastest nearest neighbour searches
# library(ncdf4) # For opening and working with NetCDF files
library(tidync, lib.loc = "../R-packages/") # For a more tidy approach to managing NetCDF data
library(SDMTools) # For finding points within polygons
library(lubridate) # For convenient date manipulation

# Set number of cores
doMC::registerDoMC(cores = 50)

# Disable scientific notation for numeric values
  # I just find it annoying
options(scipen = 999)

# Corners of the study area
NWA_corners <- readRDS("data/NWA_corners.Rda")

# Individual regions
NWA_coords <- readRDS("data/NWA_coords_cabot.Rda")

# The base map
map_base <- ggplot2::fortify(maps::map(fill = TRUE, col = "grey80", plot = FALSE)) %>%
  dplyr::rename(lon = long) %>%
  mutate(group = ifelse(lon > 180, group+9999, group),
         lon = ifelse(lon > 180, lon-360, lon)) %>% 
  select(-region, -subregion)
```

## Pixel prep

Up first we take the lon/lat grid from the 1/4 degree daily NOAA OISST product and find which points fall within each region. We will save this information to allow us to then easily pull out the desired pixels from the cube of OISST data.

```{r pixel-regions}
# Load NAPA bathymetry
# NAPA_bathy <- readRDS("data/NAPA_bathy.Rda")# %>% 
  # mutate(index = paste0(lon, lat))
OISST_grid <- data.frame(expand.grid(c(seq(0.125, 179.875, by = 0.25), seq(-179.875, -0.125, by = 0.25)),
                                        seq(-89.875, 89.875, by = 0.25)))
colnames(OISST_grid) <- c("lon", "lat")
# saveRDS(OISST_grid, "data/OISST_grid.Rda")

# Trim down OISST grid for faster processing
OISST_grid_region <- OISST_grid %>% 
  filter(lon >= min(NWA_coords$lon),
         lon <= max(NWA_coords$lon),
         lat >= min(NWA_coords$lat),
         lat <= max(NWA_coords$lat))
# saveRDS(OISST_grid_region, "data/OISST_grid_region.Rda")

# Function for finding and cleaning up points within a given region polygon
pnts_in_region <- function(region_in){
  region_sub <- NWA_coords %>% 
    filter(region == region_in)
  coords_in <- pnt.in.poly(pnts = OISST_grid_region[1:2], poly.pnts = region_sub[2:3]) %>% 
    filter(pip == 1) %>% 
    dplyr::select(-pip) %>% 
    mutate(region = region_in)
  return(coords_in)
}

# Run the function
NWA_info <- plyr::ldply(unique(NWA_coords$region), pnts_in_region)
# saveRDS(NWA_info, "data/NWA_info.Rda")

# Visualise to ensure success
ggplot(NWA_coords, aes(x = lon, y = lat)) +
  geom_polygon(aes(fill = region), alpha = 0.2) +
  geom_point(data = NWA_info, aes(colour = region)) +
  geom_polygon(data = map_base, aes(group = group), show.legend = F) +
    coord_cartesian(xlim = NWA_corners[1:2],
                  ylim = NWA_corners[3:4]) +
  labs(x = NULL, y = NULL)
```

## SST prep

With the OISST pixels successfully assigned to regions based on their thermal properties we now need to go about clumping these SST pixels into one mean time series per region.

```{r sst-clump, eval=FALSE}
# The OISST data location
OISST_files <- dir("../../data/OISST", full.names = T)

# The files with data in the study area
OISST_files_sub <- data.frame(files = OISST_files,
                              lon = c(seq(0.125, 179.875, by = 0.25), seq(-179.875, -0.125, by = 0.25))) %>% 
  filter(lon >= min(NWA_info$lon), lon <= max(NWA_info$lon)) %>% 
  mutate(files = as.character(files))

# Function for loading the individual OISST NetCDF files and subsetting SST accordingly
# file_name <- OISST_files_sub$files[1]
load_OISST_sub <- function(file_name, coords = NWA_info){
  res <- tidync(file_name) %>%
    hyper_filter(lat = dplyr::between(lat, min(coords$lat), max(coords$lat)),
                 time = dplyr::between(time, as.integer(as.Date("1993-01-01")),
                                       as.integer(as.Date("2018-12-31")))) %>%
    hyper_tibble() %>% 
    mutate(time = as.Date(time, origin = "1970-01-01")) %>% 
    dplyr::rename(temp = sst, t = time) %>% 
    select(lon, lat, t, temp) %>% 
    left_join(coords, by = c("lon", "lat")) %>% 
    filter(!is.na(region))
  # return(res)
}

# Clomp'em 
system.time(
  OISST_region <- plyr::ldply(OISST_files_sub$files,
                           .fun = load_OISST_sub,
                           .parallel = TRUE) %>% 
    group_by(region, t) %>% 
    summarise(temp = mean(temp, na.rm = T))
) # 18 seconds

# Save
# saveRDS(OISST_region, "data/OISST_region.Rda")
```

## MHW detection

With our clumped SST time series ready the last step in this vignette is to detect the MHWs within each.

```{r MHW-sub, eval=FALSE}
# Load the time series data
OISST_region <- readRDS("data/OISST_region.Rda")

# Calculate base results
system.time(
OISST_region_MHW <- OISST_region %>%
  group_by(region) %>%
  nest() %>%
  mutate(clims = map(data, ts2clm,
                     climatologyPeriod = c("1993-01-01", "2018-12-31")),
         events = map(clims, detect_event),
         cats = map(events, category)) %>%
  select(-data, -clims)
) # 2 seconds
# saveRDS(OISST_region_MHW, "data/OISST_region_MHW.Rda")
```

With the MHWs detected, let's visualise the results to ensure everything worked as expected.

```{r MHW-vis}
# Load MHW results
OISST_region_MHW <- readRDS("data/OISST_region_MHW.Rda")

# Events
OISST_MHW_event <- OISST_region_MHW %>%
  select(-cats) %>%
  unnest(events) %>%
  filter(row_number() %% 2 == 0) %>%
  unnest(events)

event_lolli_plot <- ggplot(data = OISST_MHW_event , aes(x = date_peak, y = intensity_max)) + 
        geom_lolli(colour = "salmon", colour_n = "red", n = 3) + 
  labs(x = "Peak Date", y = "Max. Intensity (°C)") +
  # scale_y_continuous(expand = c(0, 0))+
  facet_wrap(~region)
# ggsave(plot = event_lolli_plot, filename = "output/event_lolli_plot.pdf", height = 7, width = 13)

# Visualise
event_lolli_plot
```

Everything appears to check out. Up next in the [Variable preparation](https://robwschlegel.github.io/MHWNWA/var-prep.html) vignette we will go through the steps necessary to build the data that will be fed into our self-organising maps as seen in the [Self-organising map (SOM) analysis](https://robwschlegel.github.io/MHWNWA/som.html) vignette.

<!-- One last point however is that according to @Richaud2016 the different slopebreaks for the different regions occur at different depths, from 50 m to 400 m depending. They do however note that using a static definition of 200 m for the break does not produce significantly different results. Therefore, for the time being we will maintain the spatial breaks used above. -->

## References
